### 2019-06-19 2-Chapters/1-Chapter/chapter1.tex
New references. I need to finish Ch1 and Ch2

\TODOL{Mardi soir : je dois encore relire le 1.1 pour les éventuelles erreurs historiques !}
\TODOL{Mardi soir : je dois encore améliorer le 1.2 et être plus concis ?}
\TODOL{Mardi soir : je dois relire la fin du 1.3, et terminer le dernier point}

@@

% % Main problem
% \textbf{Main problems.}
%
At the root of the problems that motivates this thesis are the questions of global warming, and the increase of the world population.
% https://en.wikipedia.org/wiki/History_of_radio
During the last 150 years, mankind has developed a wide range of different communication technologies, and since the late $1890$s wireless communications between devices have been made possible, and more and more frequent in our lives.
With the avent of Internet of Things networks (IoT), billions of autonomous low-power devices are expected to be deployed world-wide, for a large range of different applications.
It is now well-known that with the current trend of population increase and with the energy crisis, any newly deployed technology should be cheap and energy efficient,
as well as adapted to serve a large number of people and devices.

@@

% Central broadcasting
Starting from FM radio music and news, then old black-and-white TV and color TV, the first systems were purely made for centralized broadcasting of information. In a large area, one antenna well located emits continuously in a fixed Radio Frequency (RF) band, and many devices can listen to this RF traffic and decode it, for instance to listen to music.
Different standards have been developed, and some of them are still in use today, and their nature requires a central authority that allocates RF bands to different usages (\eg, TV channels etc).
For example, in the city of Cesson-Sévigné where the campus of CentraleSupélec is located, a high RF tower emits on many standards, and in particular the \SI{92.3}{\mega\hertz} band is allocated to a French classical music radio (\emph{Radio Classique}).
Broadcast-based systems are not only used for TV or radio,
and for instance two other famous and world-wide deployed examples are the long-range Global Positioning System (GPS), and short-range infrared remote controllers.

@@

\textbf{The spectrum scarcity issue.}
%
A major problem of current wireless technologies is the issue of spectrum scarcity:
in most kinds of frequency bands, the entire RF spectrum is now allocated and free bands no longer exist, limiting the possibility of adding new usage.
As illustrated in Figure~\ref{fig:1:United_States_Frequency_Allocations_Chart_2016_The_Radio_Spectrum} below, only a very small portion of the RF spectrum in the USA is not yet allocated, from $0$ to \SI{9}{\kilo\hertz}, while the rest of the displayed portion of the spectrum, from \SI{9}{\kilo\hertz} to \SI{300}{\giga\hertz}, is allocated to various usages, that goes from maritime radionavigation (historically the first usage of radio telecommunication) to space research, inter-satellite, mobile telephony and a large number of other applications.
%
% Traditional regulatory structures have been built for an analog model and are not optimized for cognitive radio.
Regulatory bodies in the world, like the \href{https://www.fcc.gov/}{Federal Communications Commission} in the USA (see \href{https://www.fcc.gov/}{\texttt{FCC.gov}})
or \href{http://www.tdf.fr/}{TDF} in France (see \href{https://www.TDF.fr/}{\texttt{TDF.fr}}, previously known as \emph{``TéléDiffusion de France''}), as well as different independent measurement campaigns, found that most radio frequency spectrum was inefficiently utilized,
meaning that while a band can be allocated to a certain unique usage, it can be free from any user in certain times and/or places.
We refer to \cite{valenta2010survey} for a survey on spectrum utilization in Europe.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.65\textwidth,angle=90]{United_States_Frequency_Allocations_Chart_2016_The_Radio_Spectrum.pdf}
    \caption{A chart representing the allocation of radio spectrum in the USA in $2016$. Only a small share of the spectrum is not yet allocated, the top-left corner in white (it is in logarithmic scale from \SI{0}{\kilo\hertz} to \SI{300}{\giga\hertz}). \copyright{} Wikimedia.}
    % https://upload.wikimedia.org/wikipedia/commons/c/c7/United_States_Frequency_Allocations_Chart_2016_-_The_Radio_Spectrum.pdf?uselang=fr
    \label{fig:1:United_States_Frequency_Allocations_Chart_2016_The_Radio_Spectrum}
\end{figure}

\TODOL{Je peux améliorer (ou enlever) ce passage qui site J.Toledano, ou alors le commenter en plus de détails ? J'aime bien son rapport en fait (j'aurai du le lire au début de ma thèse !)}
% https://www.economie.gouv.fr/files/files/PDF/french-spectrum-mission-executive-summary-2014-06-25.pdf
% https://www.economie.gouv.fr/files/files/PDF/rapport-gestion-dynamique-spectre-2014-06-30.pdf
To highlight that the spectrum scarcity is not only an issue in the USA,
we quote the abstract of the ``Dynamic Spectrum Management For Innovation And Growth'' report by Joëlle Toledano \cite{Toledano2014EnglishSummary,Toledano2014FrenchFull}, who was commissioned by the French government in $2014$.
Different important problems are mentioned:
%
\begin{itemize}\tightlist
    \item
    \emph{``In France, digital economy represents $5\%$ of GDP and affects $80\%$ of the French economy.''}
    And the usage of wireless communications is nowadays world-wide and omnipresent:
    \emph{``Radio frequencies are essential to many sectors: communications, broadcasting, transportation, satellite networks, energy networks and smart grids, public or private security, defense, etc.''}
    Thus it is of highest interest to set up efficient, green, low-cost and durable new technologies for the future wireless standards, networks and devices.
    %
    \item
    The radio spectrum is not only widely used for all kinds of applications, it is nowadays intensely used for some of them.
    \emph{``Everyone agrees on the growing need for spectrum.
    This growing need results from two phenomena.
    On the one hand, mobile traffic should be multiplied by 13 to 25 fold between 2011 and 2017.
    On the other hand, the development of new innovative services such as the Internet of Things and its multiple applications (smart cities, e-health...), could lead to a growing number of connected devices, up to 50 billion in 2020, according to estimations.''}
    The first estimation was correct, and the mobile traffic is again \href{https://www.ericsson.com/en/mobility-report/future-mobile-data-usage-and-traffic-growth}{estimated} to be increased by 10 times between 2019 and 2025.
    % https://www.ericsson.com/en/mobility-report/future-mobile-data-usage-and-traffic-growth
    % https://www.cisco.com/c/en/us/solutions/collateral/service-provider/visual-networking-index-vni/white-paper-c11-738429.html
    If the second estimation was overestimating the growth of the IoT market, forecasts predict  more than 64 billion IoT devices by 2025.
    % https://www.businessinsider.com/internet-of-things-report?IR=T
    \item
    \emph{``Today, there are no more available frequencies in the easily exploitable frequency bands.
    Moreover, it is getting more and more difficult to resort to classical methods of frequency bands liberation.''}
    As illustrated above in Figure~\ref{fig:1:United_States_Frequency_Allocations_Chart_2016_The_Radio_Spectrum}.
    \item
    \emph{``Growing recourse to spectrum sharing, and in particular dynamic spectrum sharing, constitutes an important spectrum reserve.
    A particular sharing form, the use of unlicensed bands, open to all and free, has been intensely developed with the growth of Wi-Fi use.''}
    This thesis focusses on spectrum sharing in unlicensed bands, for IoT networks that lie at the opposite direction as the one underlying Wi-Fi networks: the spectrum sharing will be learned automatically by the end-devices, and not centrally decided by the base station.
\end{itemize}

@@

In this thesis, we study Internet of Things networks, and these two constraints of low power consumption (or long battery life) and low duty cycle are essential.
More precisely, we study the possible interconnections between \textbf{Machine Learning} and \textbf{Cognitive Radio} applied for IoT networks.
Let us define and detail both concepts.

@@

\TODOL{Emilie pense que je devrais faire ça mieux :
    Quant au survey du machine learning (Figure 1.1), je ne pense pas que
    cela soit nécessaire. Je trouve ça plus pertinent d'introduire le
    problème de décision séquentielle qui nous intéresse (qui n'est pas
    juste un sub-field du machine learning d'ailleurs, la communauté
    statistique s'y intéresse depuis bien plus longtemps !), sur l'exemple
    de l'OSA.
}

@@

% Modified from the https://draw.io/ map from
% https://github.com/trekhleb/homemade-machine-learning/blob/master/images/machine-learning-map.xml
\begin{figure}[h!]
    \centering
    \includegraphics[width=0.95\textwidth]{machine-learning-map.pdf}
\caption{Classification of the main types of Machine Learning: MAB is a subset of Reinforcement Learning and RL is a subset of ML. MAB is RL with partial information, and RL is ML with no supervision or data, just an environment to interact with.}
\label{fig:1:MainTypesOfMachineLearning}
\end{figure}

We illustrate this idea of a \emph{learning cycle} alternating between actions and feedback,
in the following Figure~\ref{fig:1:ReinforcementLearningCycle}.
A player (or learner) interacts with its environment by taking an action $A(t)$ (\eg, a choice in a finite set, $A(t)\in\{1,\dots,K\}$, or a vector $A(t)\in\R^d$), and then observing a reward $r(t)$, which is a certain measure of success from the environment (\eg, $r(t)\in\{0,1\}$ for binary failure/success, or $r(t)\in\R$).
The goal of the player is to maximize its reward, by trials and errors (\ie,actions and rewards).
Many real-world problems can be framed as Reinforcement Learning problems, as illustrated by the survey \cite{bouneffouf2019survey} and Section~\ref{sec:2:applicationsofStochasticMAB}, for instance learning to walk or to drive, learning to play a board or computer game, or discovering which treatment is efficient in healing a certain disease (clinical trial), etc.

@@

% Il faut expliquer ce que c'est que les bandits, rapidement, et leur application à l'OSA
More precisely, in this thesis we focus on a certain kind of Reinforcement Learning models where the learning does not have access to the entire reaction of the environment after taking its action.
In other words, we focus on RL with limited feedback, when the player only sees the reward given by its action at every round, and not the reward that would have been given if she chose any other action.
This kind of limited feedback is called \emph{bandit information}, and we discuss the history and the concept of \emph{Multi-Armed Bandits} in details in the next Chapter~\ref{chapter:2}.
Clinical trials and treatment discovery was historically the first application of MAB, in $1933$ with the early work of W. Thompson \cite{Thompson33}.
%
Multi-Armed Bandits (MAB) are a simple yet powerful example of the well-known exploration/exploitation dilemma:
when facing a set of $K$ actions whose effects on the environment are unknown, a learning must balance between
exploration of the unknown actions, in order to collect more information about them,
and exploitation of the best action according to its current knowledge.
%
The MAB problem have been studied in both the machine learning and the statistics community, since the $1950$s with pioneers like H. Robbins \cite{Robbins52}, and more recently it is an active field of research, since the late $1990$s \cite{Anantharam87a,Anantharam87b,auer1995gambling,Agrawal95}.
Research on MAB has produced a vast literature in the $2000$s \cite{Auer02,Auer02NonStochastic,Audibert2009minimax} and continues to be a topic of high interest, as illustrated by the surveys and books \cite{Bubeck12,LattimoreBanditAlgorithmsBook,Slivkins2019}.

@@

\paragraph{Chapter~\ref{chapter:25}.}
%
Because there are so many different MAB algorithms, we are also interested by the question of how practitioner can chose the one she will implement, in a given IoT object, in order to let this object adapts robustly in any environment.
To answer this question, we present two approaches.
The first one is an empirical comparison of the most efficient and well-known existing algorithms, in Section~\ref{sec:3:reviewSPAlgorithms} and \ref{sec:3:timeAndMemoryCosts}.
We confirm that famous and not too sophisticated algorithms, such as \UCB{} \cite{Auer02}, Thompson sampling \cite{Thompson33} and \klUCB{} \cite{KLUCBJournal}, are the most efficient in terms of regret, and offer a good balance between regret and (time and memory) complexity.
The second approach is an online algorithm selection, consisting in aggregating a (finite) set of algorithms and automatically discovering which one performs the most efficiently, against a given problem, with our contribution \Aggr{} that we detail in Chapter~\ref{chapter:25}.
This work on aggregating MAB algorithms was motivated for the OSA case of Cognitive Radio, where many previous research works only considered the $\UCB_1$ algorithm without justifying this choice rigorously,
and was presented at the IEEE WCNC conference in Barcelona, Spain, in April $2018$.

@@

\paragraph{Chapter~\ref{chapter:4}.}
%
In Chapter~\ref{chapter:4}, we study two models of IoT networks.
The first model considers IoT devices that simply have data to send to a base station, at a regular frequency (but random times), and that use the acknowledgement as feedback, in order to learn the best channel for their uplink communications.
The goal of this first model of applying decentralized RL on the device side is to increase the Quality of Service (QoS) of the application of such IoT network, by reducing the rate of failed transmission. The base station will thus receive more up-link packets from the devices it serves, if they can successfully learn an efficient spectrum access scheme.
%
On the other hand, the second model we study considers packet retransmissions, and while the concept stays similar, the application of an efficient decentralized learning results in both an increased battery life for each of the devices, as well as an increase QoS for the complete network.

We present in Section~\ref{sec:4:firstModel} a first model of an IoT network, that is made of thousands (or more) of end-devices and a single base station.
We consider a time-frequency slotted ALOHA-based protocol, by assuming a perfect synchronization between all devices and the base station.
%
Our IoT network model contains two types of devices: \emph{static devices} that use only one channel (fixed in time), and \emph{dynamic devices} that can choose the channel for each of their transmissions.
Static devices form an interfering traffic, which could be generated by devices using other standards as well.
Dynamic devices can use RL, and especially MAB algorithms, to optimize their sequential choices of channels, in order to maximize the number of successful up-link transmissions, thanks to the success/failure feedback that is the presence/absence of acknowledgement (\Ack) from the base station.
%
We first detail three base-lines, a naive approach and two centralized ``oracle'' approaches,
that are used to evaluate the performance of MAB algorithms (\UCB{} and TS), in terms of the centralized system-wide successful communications rate.
We show that our approach is interesting, as these low-cost MAB algorithms have near-optimal performance, even when the proportion of end-devices increases and the interfering traffic from the other devices becomes more and more fluctuating and unpredictable.
%
Even without changing anything on the level of IoT standards, our proposal is just an add-on capability that can be used on a unit-per-unit basis.
This model constitutes the first article written during this thesis with R. Bonnefoi, another PhD student of our team SCEE.
We presented this first work, which obtained the ``best paper award'', at the EAI CROWNCOM conference in Lisboa, Portugal in September $2017$ \cite{Bonnefoi17}.

@@

On the first hand, in the OSA setting, we failed to obtain positive result for the Selfish policy, as we proved that in some limited settings (\eg, $K=3$ and $M=2$), \Selfish-\UCB{} can suffer from linear regret with a small probability, and thus suffers from linear mean regret.
This result was later confirmed and generalized by \cite{BoursierPerchet18}.
%
On the other hand, we were able to propose new algorithms for this multi-player bandit problem with sensing information, and we analyzed our proposal \MCTopM-\klUCB, to show that it achieves a finite-time logarithmic regret upper bound, improving over previous state-of-the-art results.
%
Unfortunately this strong theoretical result heavily depends on the presence of sensing feedback, and as such our results are not (yet) applicable to the IoT model without sensing.

@@

Following two recent previous works \cite{LiuLeeShroff17,CaoZhenKvetonXie18}, we combine an efficient index policy (\klUCB) and an efficient change-point detection test, under the assumption of bounded rewards.
Building on very recent results on the Generalized Likelihood Ratio test (GLRT) for Gaussian and sub-Gaussian variables \cite{Maillard2018GLR}, we instead focus on bounded rewards and Bernoulli distributions.
Using the fact that bounded variables in $[0,1]$ are not only $1/4$ sub-Gaussian but also sub-Bernoulli, we prove the first finite-time guarantees for the GLRT for bounded variables.
We first show finite-time bounds on both the false alarm probability of our test, and its detection delay, under mild assumption on the lengths of the stationary sequences.
Our algorithm, denoted \GLRklUCB, is then presented in two variants, whether change-points are local (\ie, only one arm mean changes at each change-point) or global (\ie, possibly all the arm means change at a time).
%
We prove that by combining the asymptotically optimal policy \klUCB, designed for stationary problems, and our new analysis of the GLRT for bounded rewards, we obtain state-of-the-art guarantees on the regret of our proposed algorithm.
We make the same kind XXX
%
The performance of \GLRklUCB{} is also illustrated on numerical experiments on synthetic data, where it is shown to outperforms all the passively adaptive policies as well as the previous actively adaptive policies.
%
This last chapter is based on our last work, since October $2018$, and it first lead to a publication at the GRETSI conference in Lille, France, in August $2019$ \cite{Besson2019Gretsi}.
Our work also lead to the article \cite{Besson2019GLRT}, that will be partially rewritten and completed with more recent results and submitted soon to a journal (probably the Journal of Machine Learning Research) in fall $2019$.

\paragraph{Conclusion.}
%
We conclude this thesis in Chapter~\ref{chapter:conclusion}, to give some perspectives and directions about possible future works.
Most of the contributions that we present in this thesis call for more developments, on both directions of providing more mathematical analysis and implementing new proof-of-concepts using real wireless hardware and realistic Internet of Things networks.
%
The Appendix of this manuscript starts with a short overview of another contribution, in Appendix~\ref{app:2:DoublingTricks} we discuss about our work on doubling-trick

@@

% ----------------------------------------------------------------------------
\paragraph{Summary of the contributions}
\label{sec:1:summaryOfContributions}

>>>>

\TODOL{Mardi soir : je dois encore relire le 1.1 pour les éventuelles erreurs historiques !}
\TODOL{Mardi soir : je dois encore améliorer le 1.2 et 1.3 ?}

@@

% % Main problem
% \textbf{Main problems.}
%
At the root of the problems that motivates this thesis are the questions of global warming, and the increase of the world population.
% https://en.wikipedia.org/wiki/History_of_radio
During the last 150 years, mankind has developed a wide range of different communication technologies, and since the late $1890$s wireless communications between devices have been made possible, and more and more frequent in our lives.
With the avent of Internet of Things networks (IoT), billions of autonomous low-power devices are expected to be deployed worldwide, for a large range of different applications.
It is now well-known that with the current trend of population increase and with the energy crisis, any newly deployed technology should be cheap and energy efficient,
as well as adapted to serve a large number of people and devices.

@@

% Central broadcasting
Starting from FM radio music and news, then old black-and-white TV and color TV, the first systems were purely made for centralized broadcasting of information. In a large area, one antenna well located emits continuously in a fixed Radio Frequency (RF) band, and many devices can listen to this RF traffic and decode it, for instance to listen to music.
Different standards have been developed, and some of them are still in use today, and their nature requires a central authority that allocates RF bands to different usages (\eg, TV channels etc).
For example, in the city of Cesson-Sévigné where the campus of CentraleSupélec is located, a high RF tower emits on many standards, and in particular the \SI{92.3}{\mega\hertz} band is allocated to a French classical music radio (\emph{Radio Classique}).
Broadcast-based systems are not only used for TV or radio,
and for instance two other famous and worldwide deployed examples are the long-range global navigation satellite systems (such as USA's GPS or European Union's Galileo), and short-range infrared remote controllers.
% https://en.wikipedia.org/wiki/Satellite_navigation

@@

\textbf{The spectrum scarcity issue.}
%
A major problem of current wireless technologies is the issue of spectrum scarcity:
in most kinds of frequency bands, the entire RF spectrum is now allocated and free bands no longer exist, limiting the possibility of adding new usage.
As illustrated in Figure~\ref{fig:1:United_States_Frequency_Allocations_Chart_2016_The_Radio_Spectrum} below, only a very small portion of the RF spectrum in the USA is not yet allocated, from $0$ to \SI{9}{\kilo\hertz}, while the rest of the displayed portion of the spectrum, from \SI{9}{\kilo\hertz} to \SI{300}{\giga\hertz}, is allocated to various usages, that goes from maritime radionavigation (historically the first usage of radio telecommunication) to space research, inter-satellite, mobile telephony and a large number of other applications.
%
% Traditional regulatory structures have been built for an analog model and are not optimized for cognitive radio.
Regulatory bodies in the world, like the \href{https://www.fcc.gov/}{Federal Communications Commission} in the USA (see \href{https://www.fcc.gov/}{\texttt{FCC.gov}})
or \href{http://www.tdf.fr/}{TDF} in France (see \href{https://www.TDF.fr/}{\texttt{TDF.fr}}, previously known as \emph{``TéléDiffusion de France''}), as well as different independent measurement campaigns, found that most radio frequency spectrum was inefficiently utilized,
meaning that while a band can be allocated to a certain unique usage, it can be free from any user in certain times and/or places.
We refer to \cite{patil2011survey} for a survey on worldwide spectrum utilization, and to \cite{valenta2010survey} for the situation in Europe.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.65\textwidth,angle=90]{United_States_Frequency_Allocations_Chart_2016_The_Radio_Spectrum.pdf}
    \caption[A chart representing the allocation of radio spectrum in the USA in $2016$]{A chart representing the allocation of radio spectrum in the USA in $2016$. Only a small share of the spectrum is not yet allocated, the top-left corner in white (it is in logarithmic scale from \SI{0}{\kilo\hertz} to \SI{300}{\giga\hertz}). \copyright{} Wikimedia.}
    % https://upload.wikimedia.org/wikipedia/commons/c/c7/United_States_Frequency_Allocations_Chart_2016_-_The_Radio_Spectrum.pdf?uselang=fr
    \label{fig:1:United_States_Frequency_Allocations_Chart_2016_The_Radio_Spectrum}
\end{figure}

\TODOL{Je peux améliorer (ou enlever) ce passage qui site J.Toledano, ou alors le commenter en plus de détails ? J'aime bien son rapport en fait (j'aurai du le lire au début de ma thèse !)}
% https://www.economie.gouv.fr/files/files/PDF/french-spectrum-mission-executive-summary-2014-06-25.pdf
% https://www.economie.gouv.fr/files/files/PDF/rapport-gestion-dynamique-spectre-2014-06-30.pdf
To highlight that the spectrum scarcity is not only an issue in the USA,
we quote the abstract of the ``Dynamic Spectrum Management For Innovation And Growth'' report by J. Toledano \cite{Toledano2014EnglishSummary,Toledano2014FrenchFull}, who was commissioned by the French government in $2014$.
Different important problems are mentioned:
%
\begin{itemize}%\tightlist
    \item
    \emph{``In France, digital economy represents $5\%$ of GDP and affects $80\%$ of the French economy.''}
    And the usage of wireless communications is nowadays worldwide and omnipresent:
    \emph{``Radio frequencies are essential to many sectors: communications, broadcasting, transportation, satellite networks, energy networks and smart grids, public or private security, defense, etc.''}
    Thus it is of highest interest to set up efficient, green, low-cost and durable new technologies for the future wireless standards, networks and devices.
    %
    \item
    The radio spectrum is not only widely used for all kinds of applications, it is nowadays intensely used for some of them.
    \emph{``Everyone agrees on the growing need for spectrum.
    This growing need results from two phenomena.
    On the one hand, mobile traffic should be multiplied by 13 to 25 fold between 2011 and 2017.
    On the other hand, the development of new innovative services such as the Internet of Things and its multiple applications (smart cities, e-health...), could lead to a growing number of connected devices, up to 50 billion in 2020, according to estimations.''}
    The first estimation was correct, and the mobile traffic is again \href{https://www.ericsson.com/en/mobility-report/future-mobile-data-usage-and-traffic-growth}{estimated} to be increased by 10 times between 2019 and 2025.
    % https://www.ericsson.com/en/mobility-report/future-mobile-data-usage-and-traffic-growth
    % https://www.cisco.com/c/en/us/solutions/collateral/service-provider/visual-networking-index-vni/white-paper-c11-738429.html
    If the second estimation was overestimating the growth of the IoT market, forecasts predict  more than 64 billion IoT devices by 2025.
    % https://www.businessinsider.com/internet-of-things-report?IR=T
    \item
    \emph{``Today, there are no more available frequencies in the easily exploitable frequency bands.
    Moreover, it is getting more and more difficult to resort to classical methods of frequency bands liberation.''}
    As illustrated above in Figure~\ref{fig:1:United_States_Frequency_Allocations_Chart_2016_The_Radio_Spectrum}, and discussed in \cite{patil2011survey,valenta2010survey}.
    \item
    \emph{``Growing recourse to spectrum sharing, and in particular dynamic spectrum sharing, constitutes an important spectrum reserve.
    A particular sharing form, the use of unlicensed bands, open to all and free, has been intensely developed with the growth of Wi-Fi use.''}
    This thesis focusses on spectrum sharing in unlicensed bands, for IoT networks that lie at the opposite direction as the one underlying Wi-Fi networks: the spectrum sharing will be learned automatically by the end-devices, and not centrally decided by the base station.
\end{itemize}

@@

In this thesis, we study Internet of Things networks, and these two constraints of low power consumption (or long battery life) and low duty cycle are essential.
More precisely, we study the possible interconnections between \textbf{Cognitive Radio} and \textbf{Machine Learning} applied for IoT networks.
Let us define and detail both concepts.

@@

% \TODOL{Emilie pense que je devrais faire ça mieux :
%     Quant au survey du machine learning (Figure 1.1), je ne pense pas que
%     cela soit nécessaire. Je trouve ça plus pertinent d'introduire le
%     problème de décision séquentielle qui nous intéresse (qui n'est pas
%     juste un sub-field du machine learning d'ailleurs, la communauté
%     statistique s'y intéresse depuis bien plus longtemps !), sur l'exemple
%     de l'OSA.
% }

@@

% Modified from the https://draw.io/ map from
% https://github.com/trekhleb/homemade-machine-learning/blob/master/images/machine-learning-map.xml
\begin{figure}[h!]
    \centering
    \includegraphics[width=0.95\textwidth]{machine-learning-map.pdf}
\caption[Classification of the main types of Machine Learning]{Classification of the main types of Machine Learning: MAB is a subset of Reinforcement Learning and RL is a subset of ML. MAB is RL with partial information, and RL is ML with no supervision or data, just an environment to interact with.}
\label{fig:1:MainTypesOfMachineLearning}
\end{figure}

We illustrate this idea of a \emph{learning cycle} alternating between actions and feedback,
in the following Figure~\ref{fig:1:ReinforcementLearningCycle}.
A player (or learner) interacts with its environment by taking an action $A(t)$ (\eg, a choice in a finite set, $A(t)\in\{1,\dots,K\}$, or a vector $A(t)\in\R^d$), and then observing a reward $r(t)$, which is a certain measure of success from the environment (\eg, $r(t)\in\{0,1\}$ for binary failure/success, or $r(t)\in\R$).
The goal of the player is to maximize its reward, by trials and errors (\ie, actions and rewards).
Many real-world problems can be framed as Reinforcement Learning problems, as illustrated by the survey \cite{bouneffouf2019survey} and Section~\ref{sec:2:applicationsofStochasticMAB}, for instance learning to walk or to drive, learning to play a board or computer game, or discovering which treatment is efficient in healing a certain disease (clinical trial), etc.

@@

% Il faut expliquer ce que c'est que les bandits, rapidement, et leur application à l'OSA
More precisely, in this thesis we focus on a certain kind of Reinforcement Learning models where the learner does not have access to the entire reaction of the environment after taking its action.
In other words, we focus on RL with limited feedback, when the player only sees the reward given by its action at every round, and not the reward that would have been given if she chose any other action.
This kind of limited feedback is called \emph{bandit information}, and we discuss the history and the concept of \emph{Multi-Armed Bandits} in details in the next Chapter~\ref{chapter:2}.
Clinical trials and treatment discovery were historically the first applications of MAB since $1930$s, with the early work of W. Thompson \cite{Thompson33}.
%
Multi-Armed Bandits (MAB) are a simple yet powerful example of the well-known exploration/exploitation dilemma:
when facing a set of $K$ actions whose effects on the environment are unknown, a learning must balance between
exploration of the unknown actions, in order to collect more information about them,
and exploitation of the best action according to its current knowledge.
%
The MAB problem have been studied in both the machine learning and the statistics community, since the $1950$s with pioneers like H. Robbins \cite{Robbins52}, and more recently it is an active field of research, since the late $1990$s \cite{Anantharam87a,Anantharam87b,auer1995gambling,Agrawal95}.
Research on MAB has produced a vast literature in the $2000$s \cite{Auer02,Auer02NonStochastic,Audibert2009minimax} and continues to be a topic of high interest, as illustrated by the surveys and books \cite{Bubeck12,LattimoreBanditAlgorithmsBook,Slivkins2019}, and the wide range of applications of MAB in the recent years.

@@

\paragraph{Chapter~\ref{chapter:25}.}
%
Because there are so many different MAB algorithms, we are also interested by the question of how practitioner can chose the one she will implement, in a given IoT object, in order to let this object adapts robustly in any environment.
To answer this question, we present two approaches.
The first one is an empirical comparison of the most efficient and well-known existing algorithms, in Section~\ref{sec:3:reviewSPAlgorithms} and \ref{sec:3:timeAndMemoryCosts}.
We confirm that well-known and not too sophisticated algorithms, such as \UCB{} \cite{Auer02}, Thompson sampling \cite{Thompson33} and \klUCB{} \cite{KLUCBJournal}, are the most efficient in terms of regret, and offer a good balance between regret and (time and memory) complexity.
The second approach is an online algorithm selection, consisting in aggregating a (finite) set of algorithms and automatically discovering which one performs the most efficiently, against a given problem, with our contribution \Aggr{} that we detail in Chapter~\ref{chapter:25}.
This work on aggregating MAB algorithms was motivated for the OSA case of Cognitive Radio, where many previous research works only considered the $\UCB_1$ algorithm without really justifying this choice.
This contribution was presented at the IEEE WCNC conference in Barcelona, Spain, in April $2018$.

@@

\paragraph{Chapter~\ref{chapter:4}.}
%
In Chapter~\ref{chapter:4}, we study two models of IoT networks.
The first model considers IoT devices that simply have data to send to a base station, at a regular frequency (but random times), and that use the acknowledgement as feedback, in order to learn the best channel for their uplink communications.
The goal of this first model of applying decentralized RL on the device side is to increase the Quality of Service (QoS) of the application of such IoT network, by reducing the rate of failed transmission. The base station will thus receive more up-link packets from the devices it serves, if they can successfully learn an efficient spectrum access scheme.
%
The second model we study then considers packet retransmissions, and while the concept stays similar, the application of an efficient decentralized learning results in both an increased battery life for each of the devices, as well as an increase QoS for the complete network.

We present in Section~\ref{sec:4:firstModel} a first model of an IoT network, that is made of thousands (or more) of end-devices and a single base station.
We consider a time-frequency slotted ALOHA-based protocol, by assuming a perfect synchronization between all devices and the base station.
%
Our IoT network model contains two types of devices: \emph{static devices} that use only one channel (fixed in time), and \emph{dynamic devices} that can choose the channel for each of their transmissions.
Static devices form an interfering traffic, which could be generated by devices using other standards as well.
Dynamic devices can use RL, and especially MAB algorithms, to optimize their sequential choices of channels, in order to maximize the number of successful up-link transmissions, thanks to the success/failure feedback that is the presence/absence of acknowledgement (\Ack) from the base station.
%
We first detail three base-lines, a naive approach and two centralized ``oracle'' approaches,
that are used to evaluate the performance of MAB algorithms (\UCB{} and TS), in terms of the centralized system-wide successful communications rate.
We show that our approach is interesting, as these low-cost MAB algorithms have near-optimal performance, even when the proportion of end-devices increases and the interfering traffic from the other devices becomes more and more fluctuating and unpredictable.
%
Even without changing anything on the level of the considered IoT standard, our proposal is just an add-on capability, that can be set-up on each device, on a unit-per-unit basis.
This model constitutes the first article written during this thesis, and initiated a collaboration with R. Bonnefoi, another PhD student of our team SCEE.
We presented this first work at the EAI CROWNCOM conference in Lisboa, Portugal in September $2017$, and obtained the ``best paper award'' \cite{Bonnefoi17}.

@@

On the first hand, in the OSA setting, we failed to obtain positive result for the Selfish policy, as we proved that in some limited settings (\eg, $K=3$ and $M=2$), \Selfish-\UCB{} can suffer from linear regret with a small probability, and thus suffers from linear mean regret.
This result was later confirmed and generalized by \cite{BoursierPerchet18}.
%
On the other hand, we were able to propose new algorithms for this multi-player bandit problem with sensing information, and we analyzed our proposal \MCTopM-\klUCB, to show that it achieves a finite-time logarithmic regret upper bound, improving over previous state-of-the-art results.
Our algorithm also achieves a logarithmic number of collisions and arm switches, and allows a fixed group of $M$ devices to efficiently learn to use the $M$ best channels orthogonally for almost all their up-link communications.
%
Unfortunately this strong theoretical result heavily depends on the presence of sensing feedback, and as such our results are not (yet) applicable to the IoT model without sensing.

@@

Following two recent previous works \cite{LiuLeeShroff17,CaoZhenKvetonXie18}, we combine an efficient index policy (\klUCB) and an efficient change-point detection test, under the assumption of bounded rewards.
We build on very recent results on the Generalized Likelihood Ratio test (GLRT) for Gaussian and sub-Gaussian variables \cite{Maillard2018GLR}, and we instead focus on bounded rewards and Bernoulli distributions.
Using the fact that bounded variables in $[0,1]$ are not only $1/4$ sub-Gaussian but also sub-Bernoulli, we prove the first finite-time guarantees for the GLRT for bounded variables.
We first show finite-time bounds on both the false alarm probability of our test, and its detection delay, under mild assumption on the lengths of the stationary sequences.
Our algorithm, denoted \GLRklUCB, is then presented in two variants, whether change-points are local (\ie, only one arm mean changes at each change-point) or global (\ie, possibly all the arm means change at a time).
%
We prove that by combining the asymptotically optimal policy \klUCB, designed for stationary problems, and our new analysis of the GLRT for bounded rewards, we obtain state-of-the-art guarantees on the regret of our proposed algorithm.
Like our competitors, we make the some hypothesis on the length of the stationary intervals, and the best regret upper-bound is obtained when the algorithm knows before-hand the horizon and the number of change-points, but our algorithm does not need to know any other knowledge of the problem difficulty to be tuned optimally.
%
The performance of \GLRklUCB{} is also illustrated on numerical experiments on synthetic data, where it is shown to outperforms all the passively adaptive policies as well as the previous actively adaptive policies.
%
This last chapter is based on our last work, since October $2018$, and it first lead to a publication at the GRETSI conference in Lille, France, in August $2019$ \cite{Besson2019Gretsi}.
Our work also lead to the article \cite{Besson2019GLRT}, that will be partially rewritten and completed with more recent results and submitted soon to a journal (probably the Journal of Machine Learning Research) in fall $2019$.

\paragraph{Conclusion.}
%
We conclude this thesis in Chapter~\ref{chapter:conclusion}, to give some perspectives and directions about possible future works.
Most of the contributions that we present in this thesis call for more developments, on both directions of providing more mathematical analysis and implementing new proof-of-concepts using real wireless hardware and realistic Internet of Things networks.
%
This manuscript ends with the appendixes, first with a short overview of another contribution, in Appendix~\ref{app:2:DoublingTricks} we discuss about our work on doubling-trick, in order to remove the need for an algorithm to know the horizon of the bandit game before starting to play it.
The rest of the appendix contains a list of abbreviations and notations, then lists of figures, algorithms, code samples and tables.
The list of the bibliographical references used in this thesis are given at the end.

@@

% ----------------------------------------------------------------------------
\subsection{Summary of the contributions}
\label{sec:1:summaryOfContributions}
